Efficient toolkit implementing best practices for principal component analysis of population genetic data

### 2019-10-19 paper-bedpca.tex
add Celiac & POPRES

\subsection{Robust Mahalanobis distance\label{maha}}

@@

\subsection{Projecting PCs from a reference dataset\label{proj}}

@@

\subsection{Ancestry estimation\label{ancestry}}

@@

\subsection{Ancestry estimation}

[REDO WITH OUTLIER REMOVAL]

We performed ancestry estimation of the individuals in the UK Biobank using the 1000G data (Section \ref{ancestry}), discarding individuals with unknown ancestry or mixed ancestry. 
We did not infer ancestry for 15.5\% of UK Biobank that had a maximum p-value smaller than 0.05 (Figure \ref{fig:ancestry-pval}). 
More precisely, among ``White'', ``British'' and ``Irish'' ancestries, this represented respectively 20.4\%, 10.5\% and 19.8\%, while this represented between 48.5\% (``Chinese'') and 81.4\% (``Bangladeshi''). 
In other words, most of the UK Biobank individual who self reports as ``Bangladeshi'' could not be attributed to any 1000G population, although 1000G data includes some ``Bengali in Bangladesh'' population.
Only 10 people out of 401,048 were wrongly classified in super population of the 1000G; e.g.\ one Chinese was classified as European by our method (Table \ref{tab:ancestry-pred}). 
Our method is able to accurately differentiate between sub-continental populations such as differentiating between Pakistani, Bangladeshi and Chinese people (Table \ref{tab:ancestry-pred}).

% latex table generated in R 3.6.0 by xtable 1.8-4 package
% Thu Oct 17 13:58:41 2019
\begin{table}[ht]
\centering
\caption{Self-reported ancestry (top) of UKBB individuals and their matching to 1000G populations (left) by our method. See the description of 1000G populations at \url{https://www.internationalgenome.org/category/population/}.} 
\label{tab:ancestry-pred}
\begin{tabular}{|l|c|c|c|c|c|c|c|c|c|c|c|c|}
  \hline
 & British & Irish & White & Other White & Indian & Pakistani & Bangladeshi & Chinese & Other Asian & Caribbean & African & Other Black \\ 
  \hline
AFR\_ACB &  &  &  &  &  &  &  &  &  & 1635 & 510 & 27 \\ 
  AFR\_ASW &  &  &  &  &  &  &  &  &  & 488 & 23 & 14 \\ 
  AFR\_ESN &  &  &  &  &  &  &  &  &  &  & 103 &  \\ 
  AFR\_GWD &  &  &  &  &  &  &  &  &  &  & 15 &  \\ 
  AFR\_LWK &  &  &  &  &  &  &  &  &  &  & 38 &  \\ 
  AFR\_MSL &  &  &  &  &  &  &  &  &  &  & 25 &  \\ 
  AFR\_YRI &  &  &  &  &  &  &  &  &  & 4 & 433 & 4 \\ 
   \hline
AMR\_CLM & 1 &  &  & 53 &  &  &  &  &  &  &  &  \\ 
  AMR\_MXL &  &  & 1 & 3 &  &  &  &  &  &  &  &  \\ 
  AMR\_PEL &  &  &  & 1 &  &  &  &  &  &  &  &  \\ 
  AMR\_PUR &  &  &  &  &  &  &  &  &  &  &  &  \\ 
   \hline
EAS\_CDX &  &  &  &  &  &  &  & 1 &  &  &  &  \\ 
  EAS\_CHB &  &  & 1 &  &  &  &  & 408 & 11 &  &  &  \\ 
  EAS\_CHS &  &  &  &  &  &  &  & 350 & 6 &  &  &  \\ 
  EAS\_JPT &  &  &  &  &  &  &  & 2 & 24 &  &  &  \\ 
  EAS\_KHV &  &  &  &  &  &  &  & 3 & 16 &  &  &  \\ 
   \hline
EUR\_CEU & 293299 & 5483 & 327 & 4486 &  &  &  & 1 &  &  &  &  \\ 
  EUR\_FIN & 2 &  &  & 116 &  &  &  &  &  &  &  &  \\ 
  EUR\_GBR & 91862 & 4881 & 99 & 765 &  &  &  &  &  &  & 1 &  \\ 
  EUR\_IBS & 32 &  & 4 & 481 &  &  &  &  &  &  &  &  \\ 
  EUR\_TSI & 20 &  & 1 & 166 &  &  &  &  &  &  &  &  \\ 
   \hline
SAS\_BEB &  &  &  &  & 12 &  & 50 &  &  &  &  &  \\ 
  SAS\_GIH &  &  &  &  & 302 &  &  &  &  &  &  &  \\ 
  SAS\_ITU & 1 &  &  &  & 307 & 3 &  &  & 251 & 3 &  &  \\ 
  SAS\_PJL &  &  &  &  & 1452 & 712 &  &  & 82 & 1 &  &  \\ 
  SAS\_STU &  &  &  &  & 42 &  &  &  & 172 &  &  &  \\ 
   \hline
\end{tabular}
\end{table}

@@

>>>>

\subsection{Robust Mahalanobis distance \label{maha}}

@@

\subsection{Projecting PCs from a reference dataset \label{proj}}

@@

\subsection{Ancestry matching \label{ancestry}}

@@

\subsection{Ancestry matching}

[REDO WITH OUTLIER REMOVAL?]

We performed ancestry estimation of the individuals in the UK Biobank using the 1000G data (Section \ref{ancestry}), discarding individuals with unknown ancestry or mixed ancestry. 
We did not infer ancestry for 15.5\% of UK Biobank that had a maximum p-value smaller than 0.05 (Figure \ref{fig:ancestry-pval}). 
More precisely, among ``White'', ``British'' and ``Irish'' ancestries, this represented respectively 20.4\%, 10.5\% and 19.8\%, while this represented between 48.5\% (``Chinese'') and 81.4\% (``Bangladeshi''). 
In other words, most of the UK Biobank individual who self-reports as ``Bangladeshi'' could not be attributed to any 1000G population, although 1000G data includes several ``Bengali in Bangladesh'' (SAS\_BEB).
Only 10 people out of 401,048 were wrongly classified in ``super'' population of the 1000G; e.g.\ one Chinese person was classified as European by our method (Table \ref{tab:ancestry-pred}). 
Moreover, our method is able to accurately differentiate between sub-continental populations such as differentiating between Pakistani, Bangladeshi and Chinese people (Table \ref{tab:ancestry-pred}).

% latex table generated in R 3.6.0 by xtable 1.8-4 package
% Sat Oct 19 09:45:04 2019
\begin{table}[ht]
\centering
\caption{Self-reported ancestry (top) of UKBB individuals and their matching to 1000G populations (left) by our method. See the description of 1000G populations at \url{https://www.internationalgenome.org/category/population/}.} 
\label{tab:ancestry-pred}
\begin{tabular}{|l|c|c|c|c|c|c|c|c|c|c|c|c|}
  \hline
 & British & Irish & White & Other White & Indian & Pakistani & Bangladeshi & Chinese & Other Asian & Caribbean & African & Other Black \\ 
  \hline
AFR\_ACB &  &  &  &  &  &  &  &  &  & 1687 & 580 & 28 \\ 
  AFR\_ASW &  &  &  &  &  &  &  &  &  & 365 & 12 & 12 \\ 
  AFR\_ESN &  &  &  &  &  &  &  &  &  &  & 103 &  \\ 
  AFR\_GWD &  &  &  &  &  &  &  &  &  &  & 12 &  \\ 
  AFR\_LWK &  &  &  &  &  &  &  &  &  &  & 33 &  \\ 
  AFR\_MSL &  &  &  &  &  &  &  &  &  &  & 23 &  \\ 
  AFR\_YRI &  &  &  &  &  &  &  &  &  & 4 & 403 & 3 \\ 
   \hline
AMR\_CLM & 1 &  &  & 47 &  &  &  &  &  &  &  &  \\ 
  AMR\_MXL &  &  & 1 & 3 &  &  &  &  &  &  &  &  \\ 
  AMR\_PEL &  &  &  & 1 &  &  &  &  &  &  &  &  \\ 
   \hline
EAS\_CDX &  &  &  &  &  &  &  & 1 &  &  &  &  \\ 
  EAS\_CHB & 1 &  &  &  &  &  &  & 369 & 10 &  &  &  \\ 
  EAS\_CHS &  &  &  &  &  &  &  & 398 & 8 &  &  &  \\ 
  EAS\_JPT &  &  &  &  &  &  &  & 2 & 23 &  &  &  \\ 
  EAS\_KHV &  &  &  &  &  &  &  & 3 & 14 &  &  &  \\ 
   \hline
EUR\_CEU & 323808 & 6547 & 348 & 4904 &  &  &  & 1 &  &  &  &  \\ 
  EUR\_FIN & 3 &  &  & 115 &  &  &  &  &  &  &  &  \\ 
  EUR\_GBR & 62059 & 3682 & 79 & 482 &  &  &  &  &  &  & 1 &  \\ 
  EUR\_IBS & 35 &  & 5 & 494 &  &  &  &  &  &  &  &  \\ 
  EUR\_TSI & 20 &  & 1 & 156 &  &  &  &  &  &  &  &  \\ 
   \hline
SAS\_BEB &  &  &  &  & 10 &  & 41 &  &  &  &  &  \\ 
  SAS\_GIH &  &  &  &  & 292 &  &  &  &  &  &  &  \\ 
  SAS\_ITU & 1 &  &  &  & 270 & 1 &  &  & 224 & 2 &  &  \\ 
  SAS\_PJL &  &  &  &  & 1406 & 641 &  &  & 82 & 2 &  &  \\ 
  SAS\_STU &  &  &  &  & 39 &  &  &  & 198 &  &  &  \\ 
   \hline
\end{tabular}
\end{table}

@@

We also projected individuals using the OADP projection and then applied our ancestry detection technique to other datasets. 
When applied to the European individuals of the POPRES data \cite[]{nelson2008population}, we show that 56.5\% of individuals could not be matched to one of the 26 populations of the 1000G data (Table \ref{tab:ancestry-pred-popres}). 
This is particularly dramatic for all East-Europeans, whose only 11 people could be matched out of 179 (6.1\%).
Moreover, even though 1000G data include several ``Toscani in Italia'' (EUR\_TSI), we were able to match 14.2\% of Italians only.
Nevertheless, for individuals that we could match, they were all identified as European ancestry. We could also identify accurately sub-regions of Europe; e.g.\ all Spanish and Portugese individuals that we could match (73.1\%) were identified as ``Iberian Population in Spain'' (EUR\_IBS, Table \ref{tab:ancestry-pred-popres}).
Finally, we also projected and matched people from a case-control cohort for celiac disease composed of four European countries: Italy, the Netherlands, the UK and Finland \cite[]{dubois2010multiple}. 
Among UK individuals, 5397 (79.9\%) were matched to either ``Utah Residents with Northern and Western European Ancestry'' (EUR\_CEU) or ``British in England and Scotland'' (EUR\_GBR), 10 (0.2\%) to other European populations, 1 to ``Mexican Ancestry from Los Angeles USA'' (AMR\_MXL) and 1346 (19.9\%) could not be matched (Table \ref{tab:ancestry-pred-celiac}).
Among Finns, 1543 (62.4\%) were matched to ``Finnish in Finland'' (EUR\_FIN), 6 (0.3\%) to other European populations, and 922 (37.3\%) could not be matched.
Among Italians, 229 (22\%) were matched to EUR\_TSI, 15 (1.5\%) to EUR\_IBS and 795 (76.5\%) could not be matched (Table \ref{tab:ancestry-pred-celiac}).
